---
title: "Mathematical definition of the indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical definition of the indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
  
- id: dramoslopez2011
  title: Placido-based indices of corneal irregularity
  author:
  - family: Ramos-López
    given: Darío
  - family: Martínez-Finkelshtein
    given: Andrei
  - family: Castro-Luna
    given: Gracia M.
  - family: Piñero
    given: David
  - family: Alió
    given: Jorge L. 
  container-title: Optometry and Vision Science
  volume: 88
  URL: 'https://doi.org/10.1097/opx.0b013e3182279ff8'
  DOI: 10.1097/opx.0b013e3182279ff8
  issue: 10
  publisher: American Academy of Optometry
  page: 1220-1231
  type: article-journal
  issued:
    year: 2011
    month: 10
    
- id: dramoslopez2013
  title: Screening subclinical keratoconus with Placido-based corneal indices
  author:
  - family: Ramos-López
    given: Darío
  - family: Martínez-Finkelshtein
    given: Andrei
  - family: Castro-Luna
    given: Gracia M.
  - family: Burguera-Giménez
    given: Neus
  - family: Vega-Estrada
    given: Alfredo
  - family: Piñero
    given: David
  - family: Alió
    given: Jorge L. 
  container-title: Optometry and Vision Science
  volume: 90
  URL: 'https://doi.org/10.1097/opx.0b013e3182843f2a'
  DOI: 10.1097/opx.0b013e3182843f2a
  issue: 4
  publisher: American Academy of Optometry
  page: 335-343
  type: article-journal
  issued:
    year: 2013
    month: 4
    
- id: castroluna2020
  title: Robust keratoconus detection with Bayesian network classifier for Placido based corneal indices
  author:
  - family: Castro-Luna
    given: Gracia M.    
  - family: Martínez-Finkelshtein
    given: Andrei
  - family: Ramos-López
    given: Darío    
  container-title: Contact Lens and Anterior Eye
  volume: 43
  URL: 'https://doi.org/10.1016/j.clae.2019.12.006'
  DOI: 10.1016/j.clae.2019.12.006
  issue: 4
  publisher: Elsevier
  page: 366-372
  type: article-journal
  issued:
    year: 2020
    
- id: ahn2001
  title: Least-squares orthogonal distances fitting of circle, sphere, ellipse, hyperbola, and parabola
  author:
  - family: Ahn
    given: Sung Joon    
  - family: Rauh
    given: Wolfgang
  - family: Warnecke
    given: Hans-Jürgen
  container-title: Pattern Recognition
  volume: 34
  URL: 'https://doi.org/10.1016/S0031-3203(00)00152-7'
  DOI: 10.1016/S0031-3203(00)00152-7
  issue: 12
  publisher: Elsevier
  page: 2283-2303
  type: article-journal
  issued:
    year: 2001  
    month: 12  
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The indices computed by `rPACI` allow to discriminate between normal and irregular corneas. Most of them were introduced and validated with real datasets in scientific papers (@dramoslopez2011, @dramoslopez2013). An additional index, based in a naive Bayes classifier, was added and tested with a different database in @castroluna2020. In these papers, all indices demonstrated a good sensitivity for detection of keratoconus, a corneal disease.

The purpose of this vignette is to explain the mathematical definition and the meaning of the indices given by `rPACI`. They can be split into two categories: primary and combined indices. The primary indices are: $PI_1$, $PI_2$, $PI_3$, $SL$, AR_1, $AR_2$, $AR_3$, $AR_4$, $AR_5$. They all measure certain geometrical properties of the data distribution. Based on them, other combined indices are computed: $GLPI$ (a generalized linear model) and $NBI$ (naive Bayes index).

## Data structure

First, we assume a corneal dataframe is already available, which may have been loaded from a file using the function `readCornealTopography` (see [Workflow of rPACI pacakge](packageUsage.html)), simulated using `simulateData` (see [Simulating corneal datasets](simulating.html)), or by other ways (as long as it meets the dataset requirements below).

The dataset consist of three columns: x, y (with the X and Y Cartesian coordinates of data points) and ring index (1, 2, ...). The number of rows (number of data points) in the dataset depends on the data availability (`readCornealTopography` by default only reads complete rings, and `simulateData` also simulates complete rings), and the chosen number of innermost rings to use (an optional parameter for `readCornealTopography`, 15 by default).

```{r, echo=FALSE, warning=FALSE}
  setwd(paste(dirname(rstudioapi::getActiveDocumentContext()$path),"/../R", sep=""))
  source("rPACI.R")
```
```{r, eval=FALSE}
  dataset = simulateData(rings = 13)
  plot(dataset$x,dataset$y)
```
```{r, echo=FALSE}
  dataset = simulateData(rings = 13)
  par(mar=c(1.5,1.6,.1,.1), mgp=c(0,0.5,0))
  plot(dataset$x,dataset$y,pch=20,cex=0.3,cex.axis=0.5,asp=1, xlab="", ylab="") 
```

If $N$ is the number of complete rings to be used and $N_A$ is the number of points per ring (in the example, $N=13$ and $N_A=256$, its default value), the format in `dataset` is the following:

 x          | y           | ring index
-------     | -------     | ------- 
$x_1$       | $y_1$       | $1$ 
$x_2$       | $y_2$       | $1$
$x_3$       | $y_3$       | $1$
...         | ...         | ...
$x_{N_A}$   | $y_{N_A}$   | $1$
$x_{(N_A+1)}$ | $y_{(N_A+1)}$ | $2$
$x_{(N_A+2)}$ | $y_{(N_A+2)}$ | $2$
...         | ...         | ...
$x_{(2 N_A)}$ | $y_{(2 N_A)}$ | $2$
...         | ...         | ...
$x_{((N-1) \times N_A + 1)}$ | $y_{((N-1) \times N_A + 1)}$ | $N$
...         | ...         | ...
$x_{(N \times N_A)}$ | $y_{(N \times N_A)}$ | $N$

There are $N_A$ rows per ring ($N$ rings), and thus, a total of $N \times N_A$ data points (in the example, $N \times N_A = 13 \times 256 = 3328$). The dataset shall not contain missing data (this is guaranteed if using `readCornealTopography` with default parameters or `simulateData`).

Thus, we have $N \times N_A$ data points corresponding to $N$ complete rings, given in Cartesian coordinates. Let us define $R_k$, the set of points belonging to the $k$-th ring/mire (from the inside out) and $k \in \{1, \dots, N\}$ is the ring index. Then, a single data point $P_j=( x_j, y_j )$ (corresponding to the j-th row of the dataframe) belongs to $R_k$, i.e., $P_j \in R_k$ if $j \in J_k$, $J_k:=\{ n_k, n_k+1, ..., n_k+(N_A-1) \}$, with $n_k = 1+N_A(k-1)$ ($J_k$ is the set of indices $j$ which correspond to the $k$-th mire $R_k$, and $n_k$ is the first (smallest) of them).

## Primary indices

<!-- Several indices of corneal irregularity can be calculated from the raw measurements of a Placido disk corneal topographer (@dramoslopez2011, @dramoslopez2013). Many other similar indices exist -->

<!-- \citep{Alio2016, Prakash2012, Maeda1994, Rabinowitz1999, Issarti2019, Buehren2010}, --> 

<!-- but they cannot be computed using just the raw data of a Placido topographer, as they require besides altimetry, pachymetry, or curvature data, information about both surfaces of the cornea, or other parameters.  -->

<!-- The Placido indices or metrics based on the raw ring images have demonstrated their ability in diagnosing clinical diseases such as keratoconus and subclinical keratoconus, showing high accuracy (@dramoslopez2011, @dramoslopez2013, @castroluna2020). For the readers' convenience, a brief summary of those indices is reported below (more details and derivations can be found in those references). -->

The first step to calculate the indices is to compute the best-fitting circle (with center $C_k$ and (average) radius $AR_k$) each ring, with $k \in \{1, \dots, N\}$. This is done by least-squares fitting of a general circle equation to the Cartesian coordinates of data points in ring $k$ and then, the geometrical parameters (center and radius) are computed from the coefficients (see @ahn2001 and @dramoslopez2011 for more details). These centers and radii contain information about the displacement or off-centering of the mires, as well as their size and regularity. A high level of drift or variability in these parameters is a signal of corneal irregularity (@dramoslopez2011).

From the best-fitting circle centers $C_k$, $k \in \{1, \dots, N\}$, a regression line can be adjusted to their coordinates ($y$ over $x$), yielding a slope $m$ and a intercept (which is not used). This line contains information about the direction of the displacement of the centers, which can be a signal of keratoconus (more specifically, the upper-bottom or north-south asymmetry is a marker for keratoconus, see @dramoslopez2013).

One can find also the best-fitting ellipses (with centers $\tilde{C}_k$ and axis ratio $c_k$, i.e., ratio or quotient between the major and the minor axis) for each ring ($k \in \{1, \dots, N\}$). This process is also based on a least-squares fit of the general ellipse equation to the Cartesian coordinates of data points of mire $k$ (see @ahn2001 and @dramoslopez2011 for more details). A high level of variability in the axis ratio is also a signal of corneal irregularity (@dramoslopez2011).

With these quantities, four irregulariy indices labeled as $PI_1, PI_2, PI_3$ (from *Placido Irregularity indices*) and $SL$^[The index $SL$ (introduced in @dramoslopez2013) is not called $PI_4$ to prevent misunderstandings, since an index of that name was previously defined in @dramoslopez2011]:

* $PI_1$ 
$$ PI_1  = \frac{1}{N} \max_{1 \leq n,m \leq N}{ \|C_n - C_m \| } $$
This index measures the diameter of the set of centers $C_k$ (normalized by the total number of rings $N$), where $\| \cdot \|$ is the standard Euclidean norm in $\mathbb R^2$. A high value of this index shows a large displacement or off-centering between the mires.

* $PI_2$ 
$$ PI_2 = \frac{1}{N-1} \sum_{1 \leq n \leq N-1}{ \|C_{n+1} - C_{n} \| } $$
corresponds to the total length of the path connecting consecutive centers. $PI_2$ measures the accumulated drift between each pair of mires. In some cases, $PI_1$ (diamater) can be relatively small but the centers drift a lot from a mire to the next. $PI_2$ is able to capture this kind of irregularity, which is an abnormal sign.

* $PI_3$
$$ PI_3   =   \sqrt{ \frac{1}{N} \sum_{1 \leq k \leq N} { (c_k - \overline{c})^2 } }, \quad \text{where} \quad  \overline{c}=\frac{1}{N} \sum_{1 \leq k \leq N}  c_k $$
measures the variability of the axis ratios or eccentricities of the individual rings. A large value of $PI_3$ indicates a high variability in the shapes of the rings, and thus is a signal of irregularity.

* $SL$ 
$$ SL = |m| $$
is the absolute value of the slope of the best-fitting line to the coordinates of the centers $C_k$. High values of this index correspond to vertical mire displacements whereas low values correspond to horizontal displacements.

## Combined indices


## References

