---
title: "Workflow of the rPACI package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow of the rPACI package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
  
- id: dramoslopez2011
  title: Placido-based indices of corneal irregularity
  author:
  - family: Ramos-López
    given: Darío
  - family: Martínez-Finkelshtein
    given: Andrei
  - family: Castro-Luna
    given: Gracia M.
  - family: Piñero
    given: David
  - family: Alió
    given: Jorge L. 
  container-title: Optometry and Vision Science
  volume: 88
  URL: 'https://doi.org/10.1097/opx.0b013e3182279ff8'
  DOI: 10.1097/opx.0b013e3182279ff8
  issue: 10
  publisher: American Academy of Optometry
  page: 1220-1231
  type: article-journal
  issued:
    year: 2011
    month: 10
    
- id: dramoslopez2013
  title: Screening subclinical keratoconus with Placido-based corneal indices
  author:
  - family: Ramos-López
    given: Darío
  - family: Martínez-Finkelshtein
    given: Andrei
  - family: Castro-Luna
    given: Gracia M.
  - family: Burguera-Giménez
    given: Neus
  - family: Vega-Estrada
    given: Alfredo
  - family: Piñero
    given: David
  - family: Alió
    given: Jorge L. 
  container-title: Optometry and Vision Science
  volume: 90
  URL: 'https://doi.org/10.1097/opx.0b013e3182843f2a'
  DOI: 10.1097/opx.0b013e3182843f2a
  issue: 4
  publisher: American Academy of Optometry
  page: 335-343
  type: article-journal
  issued:
    year: 2013
    month: 4
    
- id: castroluna2020
  title: Robust keratoconus detection with Bayesian network classifier for Placido based corneal indices
  author:
  - family: Castro-Luna
    given: Gracia M.    
  - family: Martínez-Finkelshtein
    given: Andrei
  - family: Ramos-López
    given: Darío    
  container-title: Contact Lens and Anterior Eye
  volume: 43
  URL: 'https://doi.org/10.1016/j.clae.2019.12.006'
  DOI: 10.1016/j.clae.2019.12.006
  issue: 4
  publisher: Elsevier
  page: 366-372
  type: article-journal
  issued:
    year: 2020
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
The **rPaci** package is developed to facilitate the calculation and interpretation of several indices for detecting corneal irregularities, and especially keratoconus. These indices were introduced in @dramoslopez2011, @dramoslopez2013 and @castroluna2020, and proved to be effective in detecting keratoconus and early keratoconus.

## Package installation
**rPaci** depends on the **bnlearn** package.
To install and load the package and its dependencies, open an R session and write the following code:

```{r, eval = F}
install.packages("rPACI", dependencies = T)
library("rPACI")
```
```{r, echo = F, warning = F}
library("rPACI")
```
## Reading data
In order to begin with the analysis, the data exported by a typical commercial Placido disk topographer (formatted as described before) can be loaded using the function `readCornealTopography()`. In this example, a sample file with the corneal topography of a normal eye (included in the package as `N01.txt`) is used:

```{r, eval = T}
dataset_N = readCornealTopography(system.file("extdata","N01.txt", package="rPACI"))
```
This function returns a `data.frame` conveniently formatted to be used in the next step of the analysis.


## Computing the Placido irregularity indices

We can now compute the [Placido irregularity indices](indicesDefinition.html) of this data set by calling function `computePlacidoIndices()`, which takes the object returned by `readCornealTopography()` as its only argument:
<!-- (see this [vignette](indicesDefinition.html) for a detailed definition of the indices) -->

```{r, eval = T}
results_N = computePlacidoIndices(dataset_N)
results_N
```

The object returned by `computePlacidoIndices()` is of class `data.frame` and contains 12 columns and 1 row (because we are analyzing only one eye).
The first column of the returned `data.frame` is the diagnosis, based on the GLPI index, which can be either `Irregular cornea` (GLPI $\geq$ 70), `Suspect cornea` (30 $\leq$ GLPI <70) or `Normal cornea` (GLPI <30). The next column is the Naive Bayes Index (NBI), which ranges between 0 and 100 and can be interpreted as the probability of suffering from keratoconus. The remaining columns correspond to the other indices (see definitions in the previous sections).

## Plotting the results
The results of the analysis can be plotted using function `plotSingleCornea()`, which takes 3 arguments: `dataset`, a `data.frame` containing the corneal topography data, i.e., the object returned by `readCornealTopography()`; `PlacidoIndices`, a `data.frame` containing the computed Placido indices, i.e., the object returned by `computePlacidoIndices()`; and, optionally, `filename`, a character vector to be displayed on the plot (for instance, the filename of the corneal topography dataset).


```{r, eval = T, fig.width = 7, fig.height = 4}
  plotSingleCornea(dataset_N, results_N, filename = "N01.txt")
```

The left-hand side of the figure shows the input data, whereas the right-hand side shows two charts: the GLPI index plot, which visually indicates the value taken by this index on a colored scale of possible values, and the PI indices distribution, which shows the distribution of the GLPI, PI-1, PI-2 and PI-3 indices in a boxplot placed on a scale of possible values. The color of the charts indicate whether the indices fall within the *normal cornea* region (green color), *suspicious cornea* region (orange color), or *irregular cornea* region (red color). In the example shown, we can clearly see that the eye is diagnosed as normal.

## Using the wrapper function
Alternatively to the aforementioned 3 functions, the wrapper function `analyzeFile()` can be used instead, which takes 2 arguments: `path`, which is a character element indicating the location of the corneal topography file, and `drawplot`, which is a logical argument indicating whether the results should be plotted or not.
This function returns a `data.frame` containing the same information as the object returned by `computePlacidoIndices()` and, optionally, the plot returned by `plotSingleCornea()` if the argument 
`drawplot` is `TRUE`. 

```{r, eval = T, fig.width = 7, fig.height = 4}
  res_K = analyzeFile(system.file("extdata","K01.txt", package="rPACI"), drawplot=TRUE)
res_K
```


The results indicate that the patient's cornea is irregular. In this case, the plot shows a GLPI index of 100 and a distribution of PI indices ranging from 100 to 130.

## Analizing multiple patients simoultaneously

So far, these functions can only be used to analyze one file at a time. In order to analyze multiple files simultaneously, the function `analyzeFolder()` can be used. This function takes 4 arguments: `path`, to indicate the location of the folder containing the files; `fileExtension`, which indicates the extension of the files, is set to 'txt' by default; `individualPlots`, which is an optional logical argument indicating whether the plot for each file should be displayed or not; and `summaryPlot`, which is an optional logical argument indicating whether a summary plot of all files analyzed should be displayed or not. If the argument `summaryPlot()` is `TRUE`, then a barplot showing the absolute frequency of each possible value of diagnosis is depicted.

```{r, eval = T, fig.width = 7, fig.height = 4}
resultsAll = analyzeFolder(system.file("extdata", package="rPACI"), individualPlots = T, summaryPlot = T)  
resultsAll
```

This function returns a `data.frame` containing 13 columns and as many rows as files analyzed. The first 12 columns correspond to the diagnosis and the indices, as in the object returned by `computePlacidoIndices()`. The last column corresponds to the file name so that a specific patient can be easily found. To see the diagnosis for each analyzed file, the first and last columns can be selected:
```{r, eval = T}
resultsAll[,c(13,1)]
```
Note that the rows are sorted from *Irregular* to *Normal* cornea. In this example, four eyes are diagnosed as *irregular*, one as *suspect* and two as *normal*. This is an easy and straightforward way to check if any patient potentially suffers from keratoconus.